Classless Java

### 2016-03-21 OOPSLA16/sections/Implementation.tex
Section 3

\section{Bridging between IB and CB in Java}

%\footnote{ Rewriting libraries and
%  application in this language/model would result in more reusable
%  libraries, but it would be a taunting task.}
Creating a new language or language extension would be an
elegant way to illustrate the point of IB. However, 
significant amounts of engineering would be needed to build a practical
language, and to achieve a similar level of integration and tool support
as Java. To have an approach that can both illustrate IB
programming and be practical, we have instead implemented Classless Java. 
This section presents the Classless Java design. 
Disciplined use of Classless Java (that is avoiding class
definitions as done in Section~\ref{}), illustrates what \emph{pure} IB is like. 
However in Classless Java CB and IB programming can be mixed together, and all 
the practical convenience of using existing Java libraries, the full
Java language, and IDE's are available in Classless Java. 
The key to our implementation is the use of compilation agents, which 
are supported in Java, and allow us to rewrite the Java AST just
before compilation.

@@

\paragraph{Type preservation:}\bruno{Move to next section?}
We preferred to introduce self,client and hair coherence instead of referring to conventional
type preservation,
to better model how our approach behave in a object oriented software ecosystem with inheritance,
where only some units may be translated/expanded.
Note inheritance crucial influence in hair coherence, and how client coherence allows
for only some units to be translated/expanded.

This means that even a library designer can decide to start using IB and still
retain backward compatibility with its clients.

>>>>

\section{Bridging between IB and CB in Java}\label{sec:imp}

%\footnote{ Rewriting libraries and
%  application in this language/model would result in more reusable
%  libraries, but it would be a taunting task.}
Creating a new language or language extension would be an
elegant way to illustrate the point of IB. However, 
significant amounts of engineering would be needed to build a practical
language, and to achieve a similar level of integration and tool support
as Java. To have an approach that can both illustrate IB
programming and be practical, we have instead implemented Classless Java. 
This section presents the Classless Java design. 
Disciplined use of Classless Java (that is avoiding class
definitions as done in Section~\ref{}), illustrates what \emph{pure} IB is like. 
However in Classless Java CB and IB programming can be mixed together, and all 
the practical convenience of using existing Java libraries, the full
Java language, and IDE's are available in Classless Java. 
The key to our implementation is the use of compilation agents, which 
are supported in Java, and allow us to rewrite the Java AST just
before compilation. We discuss advantages and limitations of our approach.

@@

### 2016-03-21 OOPSLA16/sections/Translation.tex
Some more text. Table and flow chart.

\begin{figure*}[t]\label{figure:translation}
\centering
$\begin{array}{l}
\InTextDef{16em}{[\![\mixinAnn\ \QM{interface}\ \C_0\ \QM{extends}\ \Cs\ \oC \methods\ \cC ]\!]
}{
[\![\weakAnn\ \QM{interface}\ \C_0\ \QM{extends}\ \Cs\ \oC
\methods\ \methods' \cC
]\!]}\\
\InTextWith{\methods'=\otherMethod(\C_0,\methods)}\\

\InTextDef{16em}{[\![\weakAnn\ \QM{interface}\ \C_0\ \QM{extends}\ \Cs\ \oC \methods\ \cC ]\!]
}{
\QM{interface}\ \C_0\ \QM{extends}\ \Cs\ \oC
\methods\ \ofMethod(\C_0) \cC
}\\
\InTextWith{\valid(\C_0),\QM{of}\notin\dom(\C_0) }
\end{array}$
\caption{The translation functions of $\mixinAnn$ and $\weakAnn$.}
\end{figure*}

\begin{figure*}[t]\label{figure:valid}
\centering
\hspace{-.3in}
\begin{minipage}[b]{0.5\textwidth}
$\begin{array}{l}
\InTextDef{16em}{\C_0\ \QM{with#}\m\oR \C\ \QM{_val}\cR\QM;\in
\otherMethod(\C_0,\methods)}{}\\
\hspace{.25in}\isWith(\mBody(\QM{with#}\m, \C_0), \C_0),\
\QM{with#}\m\notin\dom(\methods)\\
\InTextDef{16em}{\C_0\ \QM_\m\oR \C\ \QM{_val}\cR\QM;\in
\otherMethod(\C_0,\methods)}{}\\
\hspace{.25in}\isSetter(\mBody(\QM_\m, \C_0), \C_0),\ \QM_\m\notin\dom(\methods)\\
\InTextDef{7ex}{\valid(\C_0)}{\forall \m\in\dom(\C_0),\mbox{ if }\mh\QM; =
            \mBody(\m, \C_0),}\\
            \hspace{.24in}\mbox{ one of the following cases is satisfied:}\\
\tab\tab
\isField(\method), \isWith(\method, \C_0) \mbox{ or }
\isSetter(\method,\C_0)\\
\end{array}$
\end{minipage}
\vline
\hspace{.1in}
\begin{minipage}[b]{0.43\textwidth}
$\begin{array}{l}
\InTextDef{24ex}{\isField(\C\ \m\oR\cR\QM;)}{
\mnot\ \specialName(\m)}\\
\InTextDef{24ex}{\isWith(\C'\ \QM{with#}\m \oR \C\ \x\cR\QM;, \C_0)}{}\\
\hspace{.3in}\C_0 <: \C', \mBody(\m, \C_0) = \C\ \m\oR\cR\QM;
\ \mand\ \mnot\ \specialName(\m)\\
\InTextDef{24ex}{\isSetter(\C'\ \QM_\m \oR \C\ \x\cR\QM;, \C_0)}{}\\
\hspace{.3in}\C_0 <: \C', \mBody(\m, \C_0) = \C\ \m\oR\cR\QM;
\ \mand\ \mnot\ \specialName(\m)\\
\end{array}$
\end{minipage}
\caption{The $\otherMethod$ and $\valid$ functions (left) and auxiliary functions (right).}
\end{figure*}

This section gives an overview of what $\mixinAnn$ generates in Classless Java. We provide a
formalization of Classless Java in Appendix~\ref{sec:formal}, which models the essence of
Java interfaces with default methods, including the syntax as well as the typing rules.
Since the formalized part of Classless Java does not consider casts or \Q@instanceof@,
the \Q@with@ method is not included in the
formal translation. For the same reason \Q@void@ returning setters are
not included, since they are just a minor variation over the more
interesting fluent setters, and they would require special handling
just for the conventional \Q@void@ type.

\begin{figure*}[t]\label{figure:ofmethod}
\centering
\hspace{-.2in}
\begin{minipage}[b]{0.53\textwidth}
$\begin{array}{l}
\InTextDef{5em}{\ofMethod(\C_0)}{
 \QM{static}\ \C_0\ \QM{of} \oR \C_1\ \QM_\m_1\QM,\ldots \C_n\ \QM_\m_n\cR\
\QM{\{}}\\
\tab\QM{return new}\ \C_0 \oR\cR\ \QM{\{}\\
\tab\tab \C_1\ \m_1 = \QM_\m_1\QM;\ldots \C_n\ \m_n = \QM_\m_n\QM; \\
\tab\tab
\C_1\ \m_1\oR\cR\ \QM{\{return }\ \m_1\QM{;\}}\ \ldots
\C_n\ \m_n\oR\cR\ \QM{\{return }\ \m_n\QM{;\}}\\
\tab\tab\withMethod(\C_1,\m_1,\C_0,\es_1)\ldots\withMethod(\C_n,\m_n,\C_0,\es_n)\\
\tab\tab\setterMethod(\C_1,\m_1,\C_0)\ldots\setterMethod(\C_n,\m_n,\C_0)\\
%\tab\tab\cloneMethod(\C_0,\es)\\
%\tab\tab\withMethod(\C_0)\\
\tab\QM{\};\}} \\
\InTextWith{\C_1\ \m_1\QM{();},\ldots \C_n\ \m_n\QM{();} = \fieldsFunc(\C_0)}\\
\hspace{.5in}\mbox{ and }\es_i=\m_1\QM,\ldots\QM, \m_{i-1}\QM,\QM{_val,}\m_{i+1}\QM,\ldots\QM, \m_n
\end{array}$
\end{minipage}
\vline
\hspace{.2in}
\begin{minipage}[b]{0.4\textwidth}
$\begin{array}{l}
\InTextDef{16ex}{\method\in\fieldsFunc(\C_0)}{}\\
\hspace{.3in}\isField(\method)\ \mand\
\method=\mBody(m^\method,\C_0)\\
\InTextDef{10em}{\withMethod(\C,\m,\C_0,\es)}{}\\
\hspace{.2in}\C_0\ \QM{with#}\m\oR \C\ \QM{_val}\cR\ \QM{\{}
\QM{return}\ \C_0\QM{.of(}\es\QM{);\}} \\
\InTextWith{\mBody(\QM{with#}\m,\C_0) \mbox{ having the form }\mh\QM;}\\
\InTextDef{10em}{\withMethod(\C,\m,\C_0,\es)}{\emptyset\mbox{ otherwise}}\\
\InTextDef{10em}{\setterMethod(\C,\m,\C_0)}{}\\
\hspace{.2in}\C_0\ \QM_\m\oR \C\ \QM{_val}\cR\ \QM{\{}
 \m\QM{= _val;return this;\}} \\
\InTextWith{
\mBody(\QM_\m,\C_0) \mbox{ having the form }\mh\QM;}\\
\InTextDef{10em}{\setterMethod(\C,\m,\C_0)}{\emptyset\mbox{ otherwise}}\\
%\cloneMethod:&\cloneMethod(\C_0,\es)=
%\C_0\ \QM{clone()\{return}\ \C_0\QM{.of(}\es\QM{);\}} \\
%&\mbox{iff }
%\mBody(\QM{clone},\C_0) \mbox{ is of form }\mh\QM;\\
%&\cloneMethod(\C_0,\es)=\emptyset\mbox{ otherwise}\\
\end{array}$
\end{minipage}
\caption{The generated $\QM{of}$ method (left) and auxiliary functions (right).}
\end{figure*}

@@

Figure~\ref{figure:translation} presents the translation functions. In the first function, $\mixinAnn$
injects refined methods to interface $\C_0$, while in the second function, $\weakAnn$ invokes
$\ofMethod(\C_0)$ and generates the $\QM{of}$
method for $\C_0$, if such a method does not exist in its domain, and all the abstract methods are
valid for the annotation. Figure~\ref{figure:valid} and Figure~\ref{figure:ofmethod}\haoyuan{Refs are wrong!?} present more details on the auxiliary
functions. The formal definition is available in Appendix~\ref{subsec:translation}.

@@

\subsection{Guarantees}

To understand to what extent our approach is correct (and more in general, what it means to say that a language tuning is correct) we identify three types of guarantees:
\begin{itemize}
\item \textit{Self coherence}: the generated code itself is well-typed; that is,
  type errors are not present in code the user has not written.
In our case it means that either \mixin{} produces (in a controlled way) an
understandable error, or the interface can be successfully annotated and the generated code is well-typed.
We guarantee \textit{Self coherence}.

\item \textit{Client coherence}: all the client code (for example method calls)
  that is well-typed before code generation is also well-typed after the generation.
The annotation just adds more behavior without removing any functionality.
We guarantee \textit{Client coherence}.

\item \textit{Heir coherence}: interfaces (and in general classes) inheriting the instrumented code are well-typed if they were well-typed without the instrumentation.
This would forbid adding any (default or abstract) method to the annotated interfaces, including type refinement.
\mixin  does not guarantee \textit{Heir coherence}.
Indeed consider the following:

\begin{lstlisting}
interface A { int x(); A withX(int x); }
@Obj interface B extends A {}
interface C extends B { A withX(int x); }
\end{lstlisting}

\noindent This code is correct before the translation, but \mixin would  generate in \Q@B@  a method ``\Q@B withX(int x);@''.
This would break \Q@C@. \\*
Similarly, an expression of the form ``\Q@new B(){.. A withX(int x){..}}@''
would be correct before translation, but would be ill-typed after the translation.
\end{itemize}

\noindent This means that our automatic type refinement
is a useful and convenient feature, but not transparent to the heirs of the annotated interface. They need to be aware of the annotation semantics and provide the right type while refining methods.

\subsection{Results}

@@

\begin{thm}[@ObjOf tuning]
If a given interface table $\II$ IT is OK
 where $\II$ has $\weakAnn$,
$\valid(\C^{\II})$  and $\QM{of}\notin\dom(\C^{\II})$,
then the interface table $[\![\II]\!]$ IT is OK.
\end{thm}

\begin{thm}[@Obj tuning]
If a given interface table $\II$ IT is OK
 where $\II$ has $\mixinAnn$,
$\valid(\C^{\II})$  and $\QM{of}\notin\dom(\C^{\II})$, and there is no heir of $\C^{\II}$,
then the interface table $[\![\II]\!]$ IT is OK.
\end{thm}

Informally, the theorems mean that for a client program that
typechecks before the translation is applied, if the annotated type has
no subtypes and no objects of that type are created, then type safety
is guaranteed after the successful translation.

The second
step of \mixin, namely what $\weakAnn$ does in the formalization, is
guaranteed to be type-safe for the three kinds of coherence by the $\weakAnn$  tuning theorem.
\mixin
tuning is more interesting: since \mixin does not guarantee heir
coherence, we explicitly exclude the presence of heirs. In this way
\mixin tuning guarantees only self and client coherence. The formal
theorem proofs are available in the supplementary materials.

>>>>

\begin{figure*}[t]
\centering
$\begin{array}{l}
\InTextDef{16em}{[\![\mixinAnn\ \QM{interface}\ \C_0\ \QM{extends}\ \Cs\ \oC \methods\ \cC ]\!]
}{
[\![\weakAnn\ \QM{interface}\ \C_0\ \QM{extends}\ \Cs\ \oC
\methods\ \methods' \cC
]\!]}\\
\InTextWith{\methods'=\otherMethod(\C_0,\methods)}\\

\InTextDef{16em}{[\![\weakAnn\ \QM{interface}\ \C_0\ \QM{extends}\ \Cs\ \oC \methods\ \cC ]\!]
}{
\QM{interface}\ \C_0\ \QM{extends}\ \Cs\ \oC
\methods\ \ofMethod(\C_0) \cC
}\\
\InTextWith{\valid(\C_0),\QM{of}\notin\dom(\C_0) }
\end{array}$
\caption{The translation functions of $\mixinAnn$ and $\weakAnn$.}
\label{figure:translation}
\end{figure*}

\begin{figure*}[t]
\centering
\hspace{-.3in}
\begin{minipage}[b]{0.5\textwidth}
$\begin{array}{l}
\InTextDef{16em}{\C_0\ \QM{with#}\m\oR \C\ \QM{_val}\cR\QM;\in
\otherMethod(\C_0,\methods)}{}\\
\hspace{.25in}\isWith(\mBody(\QM{with#}\m, \C_0), \C_0),\
\QM{with#}\m\notin\dom(\methods)\\
\InTextDef{16em}{\C_0\ \QM_\m\oR \C\ \QM{_val}\cR\QM;\in
\otherMethod(\C_0,\methods)}{}\\
\hspace{.25in}\isSetter(\mBody(\QM_\m, \C_0), \C_0),\ \QM_\m\notin\dom(\methods)\\
\InTextDef{7ex}{\valid(\C_0)}{\forall \m\in\dom(\C_0),\mbox{ if }\mh\QM; =
            \mBody(\m, \C_0),}\\
            \hspace{.24in}\mbox{ one of the following cases is satisfied:}\\
\tab\tab
\isField(\method), \isWith(\method, \C_0) \mbox{ or }
\isSetter(\method,\C_0)\\
\end{array}$
\end{minipage}
\vline
\hspace{.1in}
\begin{minipage}[b]{0.43\textwidth}
$\begin{array}{l}
\InTextDef{24ex}{\isField(\C\ \m\oR\cR\QM;)}{
\mnot\ \specialName(\m)}\\
\InTextDef{24ex}{\isWith(\C'\ \QM{with#}\m \oR \C\ \x\cR\QM;, \C_0)}{}\\
\hspace{.3in}\C_0 <: \C', \mBody(\m, \C_0) = \C\ \m\oR\cR\QM;
\ \mand\ \mnot\ \specialName(\m)\\
\InTextDef{24ex}{\isSetter(\C'\ \QM_\m \oR \C\ \x\cR\QM;, \C_0)}{}\\
\hspace{.3in}\C_0 <: \C', \mBody(\m, \C_0) = \C\ \m\oR\cR\QM;
\ \mand\ \mnot\ \specialName(\m)\\
\end{array}$
\end{minipage}
\caption{The $\otherMethod$ and $\valid$ functions (left) and
  auxiliary functions (right).}
\label{figure:valid}
\end{figure*}

This section gives an overview of what $\mixinAnn$ generates in
Classless Java, and what properties are guaranteed in the translation.

We provide a formalization of Classless Java in Appendix~\ref{sec:formal}, which models the essence of
Java interfaces with default methods, including the syntax as well as
the typing rules. Classless Java is just a proper subset of Java 8, so
it is easy to understand the translation presented in this section
without the syntax and typing rules of Classless Java.
Since the formalized part of Classless Java does not consider casts or \Q@instanceof@,
the \Q@with@ method is not included in the
formal translation. For the same reason \Q@void@ returning setters are
not included, since they are just a minor variation over the more
interesting fluent setters, and they would require special handling
just for the conventional \Q@void@ type.

\begin{figure*}[t]
\centering
\hspace{-.2in}
\begin{minipage}[b]{0.53\textwidth}
$\begin{array}{l}
\InTextDef{5em}{\ofMethod(\C_0)}{
 \QM{static}\ \C_0\ \QM{of} \oR \C_1\ \QM_\m_1\QM,\ldots \C_n\ \QM_\m_n\cR\
\QM{\{}}\\
\tab\QM{return new}\ \C_0 \oR\cR\ \QM{\{}\\
\tab\tab \C_1\ \m_1 = \QM_\m_1\QM;\ldots \C_n\ \m_n = \QM_\m_n\QM; \\
\tab\tab
\C_1\ \m_1\oR\cR\ \QM{\{return }\ \m_1\QM{;\}}\ \ldots
\C_n\ \m_n\oR\cR\ \QM{\{return }\ \m_n\QM{;\}}\\
\tab\tab\withMethod(\C_1,\m_1,\C_0,\es_1)\ldots\withMethod(\C_n,\m_n,\C_0,\es_n)\\
\tab\tab\setterMethod(\C_1,\m_1,\C_0)\ldots\setterMethod(\C_n,\m_n,\C_0)\\
%\tab\tab\cloneMethod(\C_0,\es)\\
%\tab\tab\withMethod(\C_0)\\
\tab\QM{\};\}} \\
\InTextWith{\C_1\ \m_1\QM{();},\ldots \C_n\ \m_n\QM{();} = \fieldsFunc(\C_0)}\\
\hspace{.5in}\mbox{ and }\es_i=\m_1\QM,\ldots\QM, \m_{i-1}\QM,\QM{_val,}\m_{i+1}\QM,\ldots\QM, \m_n
\end{array}$
\end{minipage}
\vline
\hspace{.2in}
\begin{minipage}[b]{0.4\textwidth}
$\begin{array}{l}
\InTextDef{16ex}{\method\in\fieldsFunc(\C_0)}{}\\
\hspace{.3in}\isField(\method)\ \mand\
\method=\mBody(m^\method,\C_0)\\
\InTextDef{10em}{\withMethod(\C,\m,\C_0,\es)}{}\\
\hspace{.2in}\C_0\ \QM{with#}\m\oR \C\ \QM{_val}\cR\ \QM{\{}
\QM{return}\ \C_0\QM{.of(}\es\QM{);\}} \\
\InTextWith{\mBody(\QM{with#}\m,\C_0) \mbox{ having the form }\mh\QM;}\\
\InTextDef{10em}{\withMethod(\C,\m,\C_0,\es)}{\emptyset\mbox{ otherwise}}\\
\InTextDef{10em}{\setterMethod(\C,\m,\C_0)}{}\\
\hspace{.2in}\C_0\ \QM_\m\oR \C\ \QM{_val}\cR\ \QM{\{}
 \m\QM{= _val;return this;\}} \\
\InTextWith{
\mBody(\QM_\m,\C_0) \mbox{ having the form }\mh\QM;}\\
\InTextDef{10em}{\setterMethod(\C,\m,\C_0)}{\emptyset\mbox{ otherwise}}\\
%\cloneMethod:&\cloneMethod(\C_0,\es)=
%\C_0\ \QM{clone()\{return}\ \C_0\QM{.of(}\es\QM{);\}} \\
%&\mbox{iff }
%\mBody(\QM{clone},\C_0) \mbox{ is of form }\mh\QM;\\
%&\cloneMethod(\C_0,\es)=\emptyset\mbox{ otherwise}\\
\end{array}$
\end{minipage}
\caption{The generated $\QM{of}$ method (left) and auxiliary functions
  (right).}
\label{figure:ofmethod}
\end{figure*}

@@

Figure~\ref{figure:translation} presents the translation functions. In the first function, $\mixinAnn$
injects refined methods to interface $\C_0$, while in the second function, $\weakAnn$ invokes
$\ofMethod(\C_0)$ and generates the $\QM{of}$
method for $\C_0$, if such a method does not exist in its domain, and all the abstract methods are
valid for the annotation.

Figure~\ref{figure:valid} and Figure~\ref{figure:ofmethod} present more details on the auxiliary
functions. \bruno{if the figures are mentioned here, what should
  follow is the explanation on what is one the figures. Please fix
  this Haoyuan and makes sure that everything that is on the figures
  is explained here. }
The formal definition is available in
Appendix~\ref{subsec:translation}.\bruno{what formal definition?}

@@

\bruno{I think there is explanation missing in this subsection;
  probably text which is now in the appendix. Haoyuan, please take
  care of this.}

\subsection{Results}\label{subsec:results}

Classless Java provides some guarantees regarding the generated
code. Essentially, Classless Java ensures the \emph{self coherence}
and \emph{client coherence} properties informally introduced in
Section~\ref{sec:imp}. Furthermore, we can show that 
\emph{if there are no type-refinements}, then \emph{heir coherence} also
holds. The result about heir coherence is possible to prove 
because the translation is split into two parts. In essence heir
coherence is a property of the translation of  $\weakAnn$, but not of 
\mixin.

@@

\bruno{Do we have good names for the theorems Marco? Cannot be called
  tunning now!}
\begin{thm}[@ObjOf]
If a given interface table $\II$ IT is OK
 where $\II$ has $\weakAnn$,
$\valid(\C^{\II})$  and $\QM{of}\notin\dom(\C^{\II})$,
then the interface table $[\![\II]\!]$ IT is OK.
\end{thm}

\begin{thm}[@Obj]
If a given interface table $\II$ IT is OK
 where $\II$ has $\mixinAnn$,
$\valid(\C^{\II})$  and $\QM{of}\notin\dom(\C^{\II})$, and there is no heir of $\C^{\II}$,
then the interface table $[\![\II]\!]$ IT is OK.
\end{thm}

Informally, the theorems mean that for a client program that
typechecks before the translation is applied, if the annotated type has
no subtypes and no objects of that type are created, then type safety
\bruno{of the generated code?}
is guaranteed after the successful translation.

The second
step of \mixin, namely what $\weakAnn$ does in the formalization, is
guaranteed to be type-safe for the three kinds of coherence by the $\weakAnn$ theorem.
The \mixin theorem is more interesting: since \mixin does not guarantee heir
coherence, we explicitly exclude the presence of heirs. In this way
the \mixin theorem guarantees only self and client coherence. The formal
theorem proofs are available in the supplementary
materials\bruno{point to appendix instead}.

\paragraph{Type preservation:} 
Note that we preferred to introduce self, client and heir coherence instead of referring to conventional
type preservation theorems~\cite{}. The reason is 
to better model how our approach behaves in a object oriented software ecosystem with inheritance,
where only some units may be translated/expanded.
Note that inheritance crucial influence in heir coherence, and how client coherence allows
for only some units to be translated/expanded.\bruno{I don't
  understand the point of the last sentence? Rephrase?}
This means that even a library designer can decide to start using IB
programming and still
retain backward compatibility with its clients.
